```{r, echo=FALSE}
rm(list=ls())
```

```{r setup, include=FALSE}
## Libraries
## You may have to install these libraries first 
## with the tab on bottom right panel : "Packages" ; click install
## or : install.packages("ggplot2")   - don't forget the ""
#library(ggplot2)
## library(knitr) # It is better not to load knit but make direct calls to the settings

library(formatR)
library(rmdformats)
library(tidyverse)
library(knitr)
library(kableExtra)
library(skimr)
library(readxl)
library(ggplot2)
library(dplyr)
library(plotly)
library(readxl)
library(readr)
library(patchwork)
#install.packages("plotly")

# For dev version
#install.packages("VIM")
#devtools::install_github("haozhu233/kableExtra")
library(kableExtra)
library(devtools)
library(VIM)

## Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)
## Knit options
knitr::opts_knit$set(
  width = 75
)
```


```{r Load_Data, eval=TRUE, echo=FALSE}
load("allgreD.RData")
load("allgreI.RData")
load("allgreM.RData")
```

# Literature Review

Our primary source of insight and conceptual foundation was the OECD's EPIC (Environmental Policies and Individual Behaviour Change) survey [1], conducted across three rounds (2008 to 2022), with an average of 13,000 observations per round and a minimum of 10 developed countries. This study provided insights into household decision-making and transport behaviors, which guided the selection of our parameters and segmentation approach.

Specifically, we followed EPIC's structure to incorporate socio-demographic variables like car ownership, household income, and the presence of children, as a foundation for our analysis. Additionally, we used this article as our reference for segmentation by trip characteristics (e.g., origin-destination regions, purpose). This methodology enabled us to have more homogeneous datasets to better capture the effects of the parameters of interest. Here is a sample outcome of the EPIC survey, which we tried to reproduce in the preliminary parts of our study.

![Choice of commuting mode in the EPIC survey [1]](./LitRevImg1.png){width="75%"}



# Research Statement

This study examines two main aspects related to car usage within households: what factors influence a household's decision to own a car, and what factors determine how car owners decide to use their car as their primary mode of transportation. 

\newline

We sought to understand the key drivers behind car ownership and the variables that influence car usage among those who own one.


# Data Preparing{.tabset} 
## Dataset{.tabset}
### AllgreD
```{r, echo=FALSE}
allgreD <- allgreD %>%
  select(-zoneres.y.1, -mode)
D_names <- names(allgreD)

D_labels <- c("Observation_Drawing_Number","Household_Number",
                    "Individual_Number_in_Household", "Trip_Number_for_Individual",
                    "Residential_Zone_Number","Trip_Purpose_at_Origin",
                    "Accompanied_Persons_Purpose_at_Origin","Origin_Zone_of_Trip",
                    "Departure_Time_Hour","Departure_Time_Minute",
                    "Trip_Purpose_at_Destination","Accompanied_Persons_Purpose_at_Destination",
                    "Number_of_Stops_During_Trip","Destination_Zone_Number","Arrival_Time_Hour",
                    "Arrival_Time_Minute","Trip_Duration_Declared",
                    "Number_of_Mechanized_Modes_Used","Transport_Costs_Covered",
                    "Travelled_Distance_As_Crow_Flies","Travelled_Distance_Declared","Trip_Element_Number",
                    "Walking_Time_At_Origin","Zone_At_Origin_of_Stop",
                    "Zone_At_Destination_of_Stop","Walking_Time_At_Destination",
                    "Vehicle_Number_Used","Number_of_Occupants_in_Vehicle",
                    "Parking_Location_Used","Parking_Type_Used","Parking_Duration",
                    "Highway_Usage","Public_Transport_Pass_Holder","Household_ID",
                    "Individual_ID","Trip_ID","Stop_ID","Number_of_Stops_in_Trip",
                    "Modified_Transport_Mode","Estimated_Transport_Cost",
                    "Aggregated_Transport_Mode","UN","DEST",
                    "ORIG","Simplified_Purpose_at_Origin",
                    "Simplified_Purpose_at_Destination"
)


D_missing_percent <- sapply(allgreD, function(x) {
  sum(is.na(x)) / length(x) * 100
})
 
D_variables <- data.frame(
  Label = D_labels,
  Missing_Percentage = D_missing_percent
) 

D_variables %>%
  arrange(desc(Missing_Percentage)) %>%
  kable() %>%
  kable_minimal()%>%
  kable_paper(full_width = F) %>%
  row_spec(c(1:4, 22, 26), extra_css = "color: red;") %>%
  row_spec(43:45, extra_css = "color: red;") %>%
  row_spec(6:10,28, extra_css = "color: blue;") %>%
  row_spec(13:17,28, extra_css = "color: green;")
```



### AllgreI
```{r, echo=FALSE}
I_names <- names(allgreI)

I_labels <- c("Observation_Drawing_Number","Household_Number","Individual_Number_in_Household",
              "Residential_Zone_Number","Gender","Relationship_to_Household_Reference","Age",
              "Has_Mobile_Phone","Has_Email","Has_Driving_License",
              "Last_Educational_Institution_Attended","Main_Occupation","Secondary_Occupation",
              "Socio_Professional_Category","Has_Public_Transport_Subscription",
              "Public_Transport_Subscription_Validity","Works_or_Studies_at_Home",
              "Work_or_Study_Location_Zone","Has_Access_to_Private_Vehicle",
              "General_Parking_Problems","Parking_Difficulties_at_Workplace",
              "Bicycle_Use_Frequency","Motorized_Two_Wheeler_Use_Frequency_Type_1",
              "Motorized_Two_Wheeler_Use_Frequency_Type_2","Car_Use_Frequency_as_Driver",
              "Car_Use_Frequency_as_Passenger","Urban_Transport_Use_Frequency",
              "Tramway_Use_Frequency","Other_Urban_Transport_Use_Frequency",
              "Transisere_Transport_Use_Frequency","Regional_Train_Use_Frequency",
              "Activity_on_Previous_Day","Household_ID","Individual_ID","Number_of_Trips_Made",
              "Total_Daily_Travel_Time","Unknown",
              "Aggregated_Socio_Economic_Status","Grouped_Socio_Professional_Category"
)


I_missing_percent <- sapply(allgreI, function(x) {
  sum(is.na(x)) / length(x) * 100
}) 

I_variables <- data.frame(
  Label = I_labels,
  Missing_Percentage = I_missing_percent
)


I_variables %>%
  arrange(desc(Missing_Percentage)) %>%
  kable() %>%
  kable_minimal()%>%
  kable_paper(full_width = F) %>%
  row_spec(c(1:3,7,9:17,18,19,21,26, 28:31,35), extra_css = "color: red;")
```


### AllgreM

```{r, echo=FALSE}
M_names <- names(allgreM)

M_labels <- c("Household Code","Household Size","Residence Area","Day of Travel","Housing Type",
                        "Occupancy Type","Dept of Reference SNCF Station",
                        "Postal Code of Reference SNCF Station","Has Telephone","Listed in Directory",
                        "Has Internet","Number of Cars Available","Type of Car 1","Fuel Type of Car 1",
                        "Year of Car 1","Engine Power of Car 1","Ownership Status of Car 1",
                        "Parking Location of Car 1","Parking Type of Car 1","Type of Car 2","Fuel Type of Car 2",
                        "Year of Car 2","Engine Power of Car 2","Ownership Status of Car 2",
                        "Parking Location of Car 2","Parking Type of Car 2","Type of Car 3","Fuel Type of Car 3",
                        "Year of Car 3","Engine Power of Car 3","Ownership Status of Car 3",
                        "Parking Location of Car 3","Parking Type of Car 3","Type of Car 4","Fuel Type of Car 4",
                        "Year of Car 4","Engine Power of Car 4", "Ownership Status of Car 4",
                        "Parking Location of Car 4","Parking Type of Car 4","Number of Bikes",
                        "Number of Motorcycles","Management Coefficient","Household ID","Individual ID",
                        "Trip ID","Stop ID","Number of People","Total Trips","Total Daily Travel Time"
)


M_missing_percent <- sapply(allgreM, function(x) {
  sum(is.na(x)) / length(x) * 100
}) 

M_variables <- data.frame(
  Label = M_labels,
  Missing_Percentage = M_missing_percent
)


M_variables %>%
  kable() %>%
  kable_minimal()%>%
  kable_paper(full_width = F) %>%
  row_spec(c(1:3,9:11,13,14:16,18:23,25:30,32,33:37,39,40,43,45,46), extra_css = "color: red;")

```

## Following steps made

> **Select our variables and merge the datasets**

```{r, include=FALSE}
AllgreD_selected <- c(
  "ORIG", "DEST", "D12", "UN", "ZONE_D_TRAJ", "mode_depl_ag", "zoneorig", "heuredep", "mindep",
  "zonedest", "heurearr", "minarr", "duree", "prisecharge", "D13", "NO_TRAJ",
  "ZONE_A_TRAJ", "TPS_MAP_ARV", "NAT_STAT", "ntraj","id_traj" ,"mode_V2", "Couteff",
  "durstat", "TPS_MAP_DEP", "nbmodemec", "autoroute", "mot_d_red", "mot_o_red",
  "tir", "zoneres.x.1", "id_men", "id_pers", "id_depl", "NB_OCCU", "LIEU_STAT","NUM_VEH"
)


AllgreD_trimmed <- allgreD[, AllgreD_selected]
AllgreI_selected <- c(
  "id_men","id_pers", "cspgroup", "OCCU1", "OCCU2", "age", "permis", "zonetrav",
  "dispovp", "btt", "lien","sexe", "ABO_TC", "nbd", "PBM_STAT", "csp", "statut2"
)



# Select only these variables from AllgreI
AllgreI_trimmed <- allgreI[, AllgreI_selected]

AllgreM_selected <- c(
  "jourdepl", "VP_DISPO", "NB_velo", "NB_2Rm", "nb_pers", "nbt2", "btt2",
  "TYPE_HAB", "TYPE_OCU", "POSSES1", "POSSES2", "POSSES3", "POSSES4", 
  "Gare2", "Gare5", "id_men"
)

# Select only these variables from AllgreM
AllgreM_trimmed <- allgreM[, AllgreM_selected]


#person_household_data <- AllgreI_trimmed %>%
#  left_join(AllgreM_trimmed, by = "id_men")
#merged_data <- AllgreD_trimmed %>%
#  left_join(person_household_data, by = "id_pers")



allgreD_updated <- AllgreD_trimmed %>%
  left_join(AllgreM_trimmed, by = "id_men")
# Merge variables from allgreI to the updated allgreD
merged_data <- allgreD_updated %>%
  left_join(AllgreI_trimmed, by = "id_pers")

merged_data <- merged_data %>%
  rename(
    id_men =id_men.x,
    crownTravel_Distance = D12,
    Area_at_origin_of_stop = ZONE_D_TRAJ,
    travel_mode = mode_depl_ag,
    departure_hour = heuredep,
    departure_minute = mindep,
    arrival_hour = heurearr,
    arrival_minute = minarr,
    declared_trip_duration = duree,
    covered_trip_cost = prisecharge,
    actualTravel_distance = D13,
    trip_number = NO_TRAJ,
    Area_at_destination_of_stop = ZONE_A_TRAJ,
    walk_time_at_destination = TPS_MAP_ARV,
    parking_type = NAT_STAT,
    number_of_stops = ntraj,
    cost_estimation = Couteff,
    parking_duration = durstat,
    walk_time_at_origin = TPS_MAP_DEP,
    highway_used = autoroute,
    purpose_at_destination = mot_d_red,
    purpose_at_origin = mot_o_red,
    residence_zone_number = tir,
    residence_area = zoneres.x.1,
    parking_location = LIEU_STAT,
    work_zone = zonetrav,
    car_availability = dispovp,
    daily_travel_time = btt,
    relationship_status = lien,
    public_trans_subscription = ABO_TC,
    number_of_trips = nbd,
    parking_problems = PBM_STAT,
    socio_category = csp,
    socio_category_group = cspgroup,
    employment_status = statut2,
    trip_day = jourdepl,
    number_of_cars = VP_DISPO,
    num_bikes = NB_velo,
    num_motorcycles = NB_2Rm,
    num_people = nb_pers,
    housing_type = TYPE_HAB,
    occupancy_type = TYPE_OCU,
    dept_sncf_station = Gare2,
    postal_sncf_station = Gare5,
    stop_id = id_traj
  
  )

merged_data <- select(merged_data, -id_men.y)

```


```{r, include=FALSE}
variables_to_factorize <- c(
  "ORIG", "DEST", "UN", "Area_at_origin_of_stop", "travel_mode", "zoneorig", "zonedest", "covered_trip_cost",
  "trip_number", "Area_at_destination_of_stop", "parking_type", "stop_id", "mode_V2", "highway_used",
  "residence_zone_number","residence_area", "id_men", "id_pers", "id_depl", "parking_location", "trip_day",
  "housing_type","occupancy_type", "POSSES1", "POSSES2", "POSSES3", "POSSES4", "dept_sncf_station",
  "postal_sncf_station", "socio_category_group", "OCCU1", "OCCU2", "permis", "work_zone", "car_availability",
  "relationship_status", "sexe", "public_trans_subscription", "parking_problems", "socio_category", 
  "employment_status","nbmodemec"
)

# Factorizing variables
merged_data <- merged_data %>%
  mutate(across(all_of(variables_to_factorize), as.factor))
# Check the structure of the dataset to confirm factorization
dataset <- merged_data
# Analyze proportion of missing mode_depl_ag
missing_mode_proportion <- sum(is.na(dataset$travel_mode)) / nrow(dataset) * 100
cat("Proportion of missing mode_depl_ag: ", missing_mode_proportion, "%\n")

if (missing_mode_proportion < 5) {
  # If missing proportion is small, remove rows with NA in mode_depl_ag
  dataset <- dataset %>% filter(!is.na(travel_mode))
}

# Verify NA count for mode_depl_ag
cat("Remaining NA count for mode_depl_ag: ", sum(is.na(dataset$travel_mode)), "\n")

dataset <- dataset %>%
  group_by(id_men) %>%
  mutate(Permis_mng = ifelse(any(as.numeric(as.character(permis)) == 1, na.rm = TRUE), 
                             1, 
                             min(as.numeric(as.character(permis)), na.rm = TRUE))) %>%
  ungroup()

```

> **Factorize the variables needed**
`r variables_to_factorize`

```{r, include=FALSE}
#save(merged_data, file = "merged_data.RData")
#load("merged_data.RData")
#load("dataset.Rdata")

data_allgre <- dataset



data_allgre <- data_allgre %>%
  mutate(real_travel_mode = ifelse(travel_mode == "VP", "VP", "Autre"))

data_allgre <- data_allgre %>%
  mutate(
    # Combine departure hour and minute into a time format
    departure_time = sprintf("%02d:%02d", departure_hour, departure_minute),
    
    # Combine arrival hour and minute into a time format
    arrival_time = sprintf("%02d:%02d", arrival_hour, arrival_minute)
  ) %>%
  select(-departure_hour, -departure_minute,-arrival_hour, -arrival_minute)

```

> **Create new variables**

+ *real_travel_mode = binary variable "VP", "Autre"*
The new data distribution for this binary variable is as follows: `r table(data_allgre$real_travel_mode)`

+ *Departure_time*

+ *Arrival_time*

+ *Filter out missing values from travel_mode(1258)*

```{r, include=FALSE}
summary(data_allgre)
```

```{r, include=FALSE}
data_allgre <- data_allgre %>%
  rename(Number_of_trips_of_the_person = nbt2,
         Daily_travel_time = btt2)

imputing_missing <- data_allgre %>%
  select(travel_mode,parking_duration,highway_used) %>%
  filter(!is.na(parking_duration)) %>%
  group_by(travel_mode) %>%
  summarise(non_missing_values = n())
imputing_missing
imputing_missing <- data_allgre %>%
  select(travel_mode,parking_duration,highway_used) %>%
  filter(!is.na(parking_duration) & travel_mode != "VP") %>%
  summarise(
    total_observations = n(),
    parking_duration_0 = sum(parking_duration == 0, na.rm = TRUE),
    highway_used_2 = sum(highway_used == 2, na.rm = TRUE)
  )
imputing_missing
#So we can see, The most values for these 2 columns are for VP travel mode. And for Non-Vp all values are almost leading to 0 parking_duration and non highway used. So finally we can impute all the missing values for no VP by 0 and 2 respectively
#So let's replace them
data_allgre <- data_allgre %>%
  mutate(parking_duration = ifelse(is.na(parking_duration) & travel_mode != "VP", 0,parking_duration),
         highway_used = ifelse(is.na(highway_used) & travel_mode != "VP", 2, highway_used))


#Number_of_stops

data_allgre <- data_allgre %>%
  mutate(number_of_stops = ifelse(is.na(stop_id), 0, number_of_stops),
         Area_at_origin_of_stop = ifelse(is.na(stop_id), "no stop", Area_at_origin_of_stop),
         Area_at_destination_of_stop = ifelse(is.na(stop_id), "no stop", Area_at_destination_of_stop),
         walk_time_at_destination = ifelse(is.na(stop_id), "no stop", walk_time_at_destination),
         walk_time_at_origin = ifelse(is.na(stop_id), "no stop", walk_time_at_origin),
         stop_id = ifelse(is.na(stop_id), "no stop", stop_id))

data_allgre <- data_allgre %>%
  mutate(car_at_your_disposal = ifelse(number_of_cars >= num_people, "YES", "NO"))

  

data_allgre <- data_allgre %>%
  mutate(car_availability =as.numeric(as.character(data_allgre$car_availability))) %>%
  group_by(id_men, id_pers) %>%
  mutate(
    car_availability = case_when(
      # Condition 1: Household has car_at_your_disposal = YES
      car_at_your_disposal == "YES" & real_travel_mode == "VP" & is.na(car_availability) ~ 1,
      car_at_your_disposal == "YES" & real_travel_mode == "Autre" & is.na(car_availability) ~ 2,
      
      # Condition 2: Household has car_at_your_disposal = NO
      car_at_your_disposal == "NO" & number_of_cars > 0 & real_travel_mode == "VP" & is.na(car_availability) ~ 1,
      car_at_your_disposal == "NO" & number_of_cars > 0 & real_travel_mode == "Autre" & is.na(car_availability) ~ 2,
      car_at_your_disposal == "NO" & number_of_cars == 0 & is.na(car_availability) ~ 3,
      
      # Retain existing non-NA values
      TRUE ~ car_availability
    )
  ) %>%
  ungroup()
```

> **Imputing missing values**

+  *`number_of_stops`, `Area_at_origin_of_stop`,`Area_at_destination_of_stop`,`walk_time_at_destination`,`walk_time_at_origin`,`stop_id`*. If there is no Stop => We can replace missing values with `no stop`

+ *`parking_duration` and `highway_used`* => If the person didn't travel by car ==> No highway used, no parking used

+ *`car_availability`* => We used the travel_mode, number_of_cars, num_people to impute those values

```{r, include=FALSE}
missing_percent <- sapply(data_allgre, function(x) {
  sum(is.na(x)) / length(x) * 100
}) 

final_variables <- data.frame(
  Missing_Percentage = missing_percent
)


final_variables %>%
  arrange(desc(Missing_Percentage)) %>%
  kable() %>%
  kable_minimal()%>%
  kable_paper(full_width = F)
```

## Deal with trip loops{.tabset}

```{r, echo=FALSE}
trip_loop <- data_allgre %>%
  group_by(id_pers) %>%
  summarise(unique_days = length(unique(trip_day)))

trip_analysis <- data_allgre %>%
  group_by(id_pers) %>%
  arrange(id_pers, departure_time) %>%  # Sort by person and departure time
  mutate(
    trip_number = row_number(),  # Assign trip sequence number
    total_trips = n()            # Total trips for each person
  ) %>%
  summarise(
    total_trips = max(total_trips),                      # Total trips for this person
    purpose_sequence = paste(purpose_at_destination, collapse = " -> "),  # Combine purposes
    mode_sequence = paste(travel_mode, collapse = "->")
  ) 


tail(trip_analysis) %>% kable() %>% kable_minimal()
```

### Old data

```{r, echo=FALSE}
filtered_data <- data_allgre %>%
  group_by(id_pers) %>%  # Group by person
  mutate(
    mode_switch = (travel_mode != lag(travel_mode, default = first(travel_mode))),  # Detect mode switch
    mode_group = cumsum(mode_switch)  # Create a unique group for each mode sequence
  ) %>%
  group_by(id_pers, mode_group) %>%  # Group by person and mode sequence
  mutate(
    trip_sequence = row_number(),  # Sequence of trips within the mode group
    total_trips_same_mode = n(),  # Total trips in the mode group
    total_crownTravel_distance = sum(crownTravel_Distance, na.rm = TRUE),  # Sum of distances (crown)
    total_actualTravel_distance = sum(actualTravel_distance, na.rm = TRUE),  # Sum of distances (actual)
    total_declared_trip_duration = sum(declared_trip_duration, na.rm = TRUE),  # Sum of durations
    end_arrival_time = max(arrival_time, na.rm = TRUE)
  ) %>%
  ungroup()



new_data <- filtered_data %>%
  filter(id_pers == "101019001") %>%
  select(id_pers, departure_time, travel_mode, mode_switch, mode_group, trip_sequence, total_trips_same_mode,total_crownTravel_distance,total_actualTravel_distance,total_declared_trip_duration,departure_time, arrival_time, end_arrival_time)
new_data %>% kable() %>% kable_styling()

```

### New data
```{r, echo=FALSE}

filtered_data <- filtered_data%>%
  filter(trip_sequence == 1) %>%  # Keep only the first trip in the sequence
  ungroup()  # Remove grouping

new_data <- filtered_data %>%
  filter(id_pers == "101019001") %>%
  select(id_pers, travel_mode, mode_switch, mode_group, trip_sequence, total_trips_same_mode,total_crownTravel_distance,total_actualTravel_distance,total_declared_trip_duration,departure_time, arrival_time, end_arrival_time)
new_data %>% kable() %>% kable_styling()
```



### Total Trips with Same Mode

```{r, echo=FALSE}
trips_number <- filtered_data %>%
  group_by(total_trips_same_mode) %>%
  summarise(
    car_users = sum(real_travel_mode == "VP"),  # Count car users
    non_car_users = sum(real_travel_mode == "Autre"),  # Count non-car users
    percentage_car_users = (car_users / (car_users + non_car_users)) * 100  # Calculate percentage
  )


ggplot(trips_number, aes(x = total_trips_same_mode, y = percentage_car_users)) +
  geom_line(color = "blue", size = 1) +  # Line plot
  geom_point(color = "red", size = 2) +  # Points for each data
  labs(
    title = "Percentage of Car Users by Total Trips with Same Mode",
    x = "Total Trips with Same Mode",
    y = "Percentage of Car Users (%)"
  ) +
  scale_x_continuous(breaks = 1:16) +  # Set x-axis breaks from 1 to 16
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

filtered_data <- filtered_data %>%
  filter(total_trips_same_mode <= 13 )
```

The analysis reveals a strong positive relationship between the number of consecutive trips made with the same mode and the percentage of car users. Specifically, when individuals anticipate making multiple trips consecutively, the likelihood of choosing a car increases significantly. The percentage of car users rises from 38% for a single trip to its peak value as the total trips made with the same mode approach 16. This trend underscores the preference for cars in scenarios involving repeated, consistent travel.

\newline

We then filtered out trips that have more than 13 consecutive trips using the car, because these probably indicates for taxi drivers that we don't want to consider in our analysis

