# Factors Influencing Household Car Ownership{.tabset}

In this analysis, we examined several key factors influencing car ownership using variables from a research paper on the same topic. These factors include the number of children and adults in the household, urban area, income, and whether the individual lives alone

\newline

To do so, we created several new variables: `Age_group`, categorized as teenagers, young adults, adults, and seniors based on the age variable; `Urban_area`, defined as major city (for Grenoble or Voiron), suburb (for St Marcellin or La Touvet), and rural (for locations outside any city); and `Income`, derived from the combination of the csp variable (representing profession) and occu1 and occu2 (indicating employment status such as full-time, part-time, or student). These new variables were essential in understanding the factors influencing car ownership.

## % Diffrence table

```{r, include=FALSE}
#Adding the purpose category
filtered_data <- filtered_data %>%
  mutate(
    trip_category = case_when(
      purpose_at_destination %in% c("TRAVAIL", "Professionnel") ~ "Work",
      purpose_at_destination %in% c("Ecole Primaire", "C.E.S.-LYCEE") ~ "Child Education",
      purpose_at_destination %in% c("UNIVERSITE") ~ "University",
      purpose_at_destination %in% c("ACHAT", "LOISIR", "ACCOMPAGNEMENT", "AUTRES") ~ "Leisure",
      purpose_at_destination == "DOMICILE" ~ "Home"
    )
  )

#Addint the age group
filtered_data <- filtered_data %>%
  mutate(
    age_group = case_when(
      age > 0 & age < 18 ~ "Teenagers",
      age >= 18 & age < 35 ~ "Young Adults",
      age >= 35 & age < 55 ~ "Adults",
      age >= 55 & age < 100 ~ "Seniors"
    ))

#Adding the income
filtered_data <- filtered_data %>%
  mutate(
    work_status_group = case_when(
      OCCU1 == 1 ~ "Full-Time Employment",  # OCCU1 == 1 → Full-Time Employment
      OCCU1 %in% c(2, 3) | (OCCU1 %in% c(4, 5, 6, 7, 8) & OCCU2 == 1) ~ "Part-Time Employment",  # Grouped part-time employment
      OCCU1 %in% c(4, 5, 6, 7, 8) & OCCU2 != 1 ~ "Unemployed",  # OCCU1 in [4, 5, 6, 7, 8] and OCCU2 != 1 → Unemployed
      OCCU1 == 9 ~ "Other",  # OCCU1 == 9 → Other
      TRUE ~ "Unknown"  # Catch-all for any undefined conditions
    )
  )




# Define the occupation codes for each category
filtered_data <- filtered_data %>%
  mutate(
    work_group = case_when(
      socio_category %in% c(21, 22, 23, 31, 32, 36) ~ "High Income",  
      socio_category %in% c(10,41, 46, 47, 48, 51, 54, 55, 56, 99) ~ "Average Income",
      socio_category %in% c(61, 66, 69, 81, 82) ~ "Low Income",
      socio_category %in% c(83, 84, 85, 86, 87, 90) ~ "No Income",
      TRUE ~ "Average Income Jobs"  # All other categories fall under Average income jobs
    )
  )
table(filtered_data$work_group, filtered_data$work_status_group)
filtered_data$income <- with(filtered_data, 
  ifelse(
    # High Income Criteria
    work_group == "High Income" & work_status_group == "Full-Time Employment", "High Income",
    
    ifelse(
      # Medium Income Criteria
      (work_group == "High Income" & work_status_group == "Part-Time Employment") |
      (work_group == "Average Income" & work_status_group %in% c("Full-Time Employment", "Other")) |
      (work_group == "Low Income" & work_status_group == "Full-Time Employment"), "Medium Income",
      
      ifelse(
        # Low Income Criteria
        (work_group == "High Income" & work_status_group == "Other") |
        (work_group == "Average Income" & work_status_group == "Part-Time Employment") |
        (work_group == "Low Income" & work_status_group %in% c("Part-Time Employment", "Other")) |
        (work_group %in% c("High Income", "Average Income", "Low Income") & work_status_group == "Unemployed"), "Low Income",
        
        # No Income Criteria
        "No Income"
      )
    )
  )
)

filtered_data <- filtered_data %>%
  mutate(income_value = case_when(
    work_group == "No Income" ~ 0,
    work_group == "Low Income" ~ 10,
    work_group == "Average Income" ~ 50,
    work_group == "High Income" ~ 80,
  ))
table(filtered_data$income_value)
#Checking when families own a car
family_possessions <- filtered_data %>%
  group_by(id_men) %>%
  summarise(
    living_alone = ifelse(n_distinct(id_pers) == 1, 1, 0),
    have_car = as.integer(any(car_availability < 3)),  # Check if any car exists in the household
    have_alternatives = as.integer(any(num_motorcycles >= 1 | num_bikes >= 1)),
    residence_area = unique(residence_area),
    number_of_children = n_distinct(id_pers[age_group == "Teenagers"]),  # Counting distinct teenagers
    number_of_adults = n_distinct(id_pers[age_group != "Teenagers"]),
    number_of_adults_group = case_when(  # Create a new column to group adults
      number_of_adults <= 1  ~ "One adult",
      number_of_adults == 2 ~ "Two adults",
      number_of_adults > 2 ~ "Three or more adults"
    ),
    numbers_of_members = number_of_children+number_of_adults,
    income_value = (sum(income_value[!duplicated(id_pers)], na.rm = TRUE)/numbers_of_members)
    ) %>%
  mutate(
    number_of_children = case_when(
      number_of_children == 0 ~ "No Children",
      number_of_children == 1 ~ "One Child",
      number_of_children == 2 ~ "Two Children",
      number_of_children > 2 ~ "Three or more Children"
    ),
    income_value = case_when(
      income_value == 0 ~"No Income",
      income_value <= 30 ~ "Low income",
      income_value <= 60 ~ "Medium income",
      income_value <= 100 ~ "High income"),
    urban_area = case_when(
      residence_area %in% c("inside_grenoble", "inside_voiron") ~ "Major city",
      residence_area %in% c("inside_la_touvet", "inside_St_Marcellin") ~ "Suburb", 
      residence_area == "outside_the_city" ~ "Rural"
    )) %>%
  select(-residence_area)


#Difference for living alone
car_ownership_alone <- family_possessions %>%
  group_by(living_alone) %>%
  summarise(
    nb_row = n(),
    car_ownership_rate = mean(have_car) * 100  # Calculate the percentage of households with cars
  ) %>%
  arrange(car_ownership_rate)

car_ownership_alone <- car_ownership_alone %>%
  mutate(
    percent_change_from_benchmark = paste(
      round(car_ownership_rate - car_ownership_rate[living_alone == 0], 2), "%")  # Compare to the 0 children group
  )


#% Difference for childrens
car_ownership_by_children <- family_possessions %>%
  group_by(number_of_children) %>%
  summarise(
    nb_row = n(),
    car_ownership_rate = mean(have_car) * 100  # Calculate the percentage of households with cars
  ) %>%
  arrange(car_ownership_rate)

car_ownership_by_children <- car_ownership_by_children %>%
  mutate(
    percent_change_from_benchmark = paste("+", 
      round(car_ownership_rate - car_ownership_rate[number_of_children == "No Children"], 2), "%")  # Compare to the 0 children group
  )

#% difference for adults
car_ownership_by_adult <- family_possessions %>%
  group_by(number_of_adults_group) %>%
  summarise(
    nb_row = n(),
    car_ownership_rate = mean(have_car) * 100  # Calculate the percentage of households with cars
  ) %>%
  arrange(number_of_adults_group)


# Calculate the percentage difference in car ownership from the benchmark (0 children)
# Using the first row (0 children) as the reference
car_ownership_by_adult <- car_ownership_by_adult %>%
  mutate(
    percent_change_from_benchmark = paste("+", car_ownership_rate - car_ownership_rate[car_ownership_by_adult$number_of_adults_group == "One adult"], "%")  # Compare to the 0 children group
  )

#Difference for residence area

car_ownership_by_area <- family_possessions %>%
  group_by(urban_area) %>%
  summarise(
    nb_rows = n(),
    car_ownership_rate = mean(have_car) * 100  # Calculate the percentage of households with cars
  ) %>%
  arrange(car_ownership_rate)



car_ownership_by_area <- car_ownership_by_area %>%
  mutate(
    urban_area = urban_area,
    percent_change_from_benchmark = paste("+", car_ownership_rate - car_ownership_rate[car_ownership_by_area$urban_area == "Major city"], "%")  # Compare to the 0 children group
  )

#Difference for income
car_ownership_by_income <- family_possessions %>%
  group_by(income_value) %>%
  summarise(
    nb_row = n(),
    car_ownership_rate = mean(have_car) * 100  # Calculate the percentage of households with cars
  ) %>%
  arrange(income_value)



car_ownership_by_income <- car_ownership_by_income %>%
  mutate(
    income_value = income_value,
    percent_change_from_benchmark = paste("+", car_ownership_rate - car_ownership_rate[car_ownership_by_income$income_value == "No Income"], "%")  # Compare to the 0 children group
  )



```

```{r, echo=FALSE}
variables_to_factorize <- c("living_alone", "have_car", "have_alternatives","number_of_children","number_of_adults_group","income_value","urban_area")

family_possessions <- family_possessions %>%
  mutate(across(all_of(variables_to_factorize), as.factor))
chi_test_alone <- chisq.test(table(family_possessions$living_alone, family_possessions$have_car))
 
p_value_alone <- sprintf("%.2e", chi_test_alone$p.value)

car_ownership_alone <- car_ownership_alone %>%
  mutate(
    category = "Living alone(Benchmark: No)",
    group = as.character(living_alone),  # Convert to character
    chisq.test_p_value = p_value_alone  # Add the p-value to all rows
  ) %>%
  select(category, group, percent_change_from_benchmark, chisq.test_p_value)  %>%
  filter(row_number() != 2)

chi_test_children <- chisq.test(table(family_possessions$number_of_children, family_possessions$have_car))
 
p_value_children <- sprintf("%.2e", chi_test_children$p.value)

car_ownership_by_children <- car_ownership_by_children %>%
  mutate(
    category = "Number of Children(Benchmark: No Children in the household)",
    group = as.character(number_of_children),
    chisq.test_p_value = p_value_children
  ) %>%
  select(category, group, percent_change_from_benchmark,chisq.test_p_value) %>%
  filter(row_number() != 1)

chi_test_adult <- chisq.test(table(family_possessions$number_of_adults_group, family_possessions$have_car))
 
p_value_adult <- sprintf("%.2e", chi_test_adult$p.value)

car_ownership_by_adult <- car_ownership_by_adult %>%
  mutate(
    category = "Number of Adults(Benchmark: 1 adult in the household",
    group = as.character(number_of_adults_group),  # Convert to character
    chisq.test_p_value = p_value_adult
  ) %>%
  select(category, group, percent_change_from_benchmark,chisq.test_p_value) %>%
  filter(row_number() !=1)

chi_test_area <- chisq.test(table(family_possessions$urban_area, family_possessions$have_car))
 
p_value_area <- sprintf("%.2e", chi_test_area$p.value)
car_ownership_by_area <- car_ownership_by_area %>%
  mutate(
    category = "Urban Area(Benchmark Major City)",
    group = as.character(urban_area),  # Convert to character
    chisq.test_p_value = p_value_area
  ) %>%
  select(category, group, percent_change_from_benchmark,chisq.test_p_value) %>%
  filter(row_number() !=1)

chi_test_income<- chisq.test(table(family_possessions$income_value, family_possessions$have_car))
 
p_value_income <- sprintf("%.2e", chi_test_income$p.value)
car_ownership_by_income <- car_ownership_by_income %>%
  mutate(
    category = "Income Group(Benchmark : No income)",
    group = as.character(income_value),  # Convert to character
    chisq.test_p_value = p_value_income
  ) %>%
  select(category, group, percent_change_from_benchmark,chisq.test_p_value) %>%
  filter(row_number() !=4)

# Combine all tables into one
car_ownership_summary <- bind_rows(
  car_ownership_by_children,
  car_ownership_by_adult,
  car_ownership_by_area,
  car_ownership_by_income,
  car_ownership_alone
)

# View the combined table
car_ownership_summary %>% 
  kable() %>%
  kable_styling() %>%
  row_spec(c(8:10), extra_css = "color: blue;") %>%
  row_spec(c(4:5), extra_css = "color: red;") %>%
  row_spec(c(6:7), 28,extra_css = "color: green;") %>%
  row_spec(11,28, extra_css = "color: orange;")
```

The data shows a clear trend: as the number of children increases, the likelihood of car ownership rises, with a 9.5% increase for one child, 11.8% for two, and 13.9% for three or more. Similarly, the percentage of households owning a car increases by 21% as the number of adults grows. Urban area also plays a role, with car ownership higher in suburban (9.65%) and rural (14.3%) areas compared to major cities. Income follows the same pattern: higher household income correlates with increased car ownership. Additionally, individuals living alone are less likely to own a car. Importantly, the chi-square test p-value confirms that there is a statistically significant association between these variables and whether a household owns a car, reinforcing the observed patterns.


## Predictive Analysis of Car Ownership{.tabset}

+ **First Model**
```{r, echo=FALSE}
library(caret)
library(rpart.plot)
library(kernlab)
library(cowplot)
library(randomForest)

library(performanceEstimation) # for SMOTE




index <- createDataPartition(family_possessions$have_car, p = 0.8, list = FALSE)
X_train <- family_possessions[index, ]
X_test <- family_possessions[-index, ]

cat("Train Data:")
table(X_train$have_car)

```


We can see that the data is imbalanced. To solve this issue we will perform the SMOTE technique:


```{r, echo=FALSE}
smote_train <- smote(have_car ~ ., data = X_train)
cat("Smote Data:")
table(smote_train$have_car)


model <- randomForest(have_car ~ living_alone  +number_of_adults_group + + have_alternatives + number_of_children+ income_value +urban_area, data = smote_train,  importance = TRUE)
model
importance_scores <- model$importance
importance_df <- as.data.frame(importance_scores)

# Add variable names as a column
importance_df$Variable <- rownames(importance_df)
importance_df <- importance_df[order(importance_df$MeanDecreaseGini, decreasing = TRUE), ]

# Plot variable importance using ggplot2
ggplot(importance_df, aes(x = reorder(Variable, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() + # Flip axes for better readability
  theme_minimal() +
  labs(
    title = "Variable Importance from Random Forest Model",
    x = "Variables",
    y = "Importance Score"
  )



```


The random forest model's out-of-bag (OBB) error rate is `r model$err.rate[500, "OOB"]*100` %, meaning approximately `r 100-model$err.rate[500, "OOB"]*100`% of predictions were correctly classified. Upon analyzing the variable importance, we found the four most important factors are `number_of_adults`, followed by the `income` in the household, whether there is only 1 individual `living_alone` and `urban area`. To optimize the model, we can try to remove the least important variables.


+ **Confusion Matrix Metrics**

```{r, echo=FALSE}
predictions <- predict(model, X_test)


accuracy_lr <- mean(predictions == X_test$have_car)
cat("Accuracy of the model: " , accuracy_lr)

confusion_matrix <- confusionMatrix(predictions, X_test$have_car)

cat("Confusion Matrix: Test Set:\n", capture.output(print(confusion_matrix)), sep = "\n")
```
The model shows an accuracy of `r confusion_matrix$overall['Accuracy']*100`, but with a significant imbalance in the class distribution, as the No Information Rate (NIR) is much higher at `r confusion_matrix$overall['No Information Rate']*100`, indicating that predicting the majority class without any model would be more accurate. The sensitivity for Class 0 is `r confusion_matrix$byClass['Sensitivity'][1]*100`, which is good, but the positive predictive value (PPV) is quite low at `r confusion_matrix$byClass['Pos Pred Value'][1]*100`, suggesting that many of the positive predictions for Class 0 are incorrect. The model's balanced accuracy of  `r confusion_matrix$overall['Accuracy']*100`` indicates that improvements are needed, especially in predicting Class 0, potentially through techniques like class balancing or model adjustments.
